---
title: "chapter5"
output: html_document
date: "2023-11-30"
editor_options: 
  markdown: 
    wrap: sentence
---

# Chapter 5

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Move the country names to rownames (see Exercise 5.5).
Show a graphical overview of the data and show summaries of the variables in the data.
Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them.
(0-3 points)

### Data import and wrangling

```{r}

library(tibble)
library(readr)
library(corrplot)
library(ggplot2)
library(tidyr)
library(GGally)

human_new <- read_csv("/Users/margot/Desktop/Desktop - MacBook Pro de MARGOT/Open data with R 2023/IODS-project/human_new.csv", show_col_types = FALSE)

# no NA
any(is.na(human_new))

# analyze and summary of variables
str(human_new)
summary(human_new)

# Country is as rownames
human_new <- column_to_rownames(human_new, "Country")

```

### Graphical overview

```{r}
# histogram per variable
pivot_longer(human_new, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + 
  geom_histogram()

# charts and correlations
ggpairs(human_new,lower=list(combo=wrap("facethist",binwidth=0.1)))

# correlation between variables
c <- cor(human_new)
corrplot(c, method="circle")
```

### Commenting distributions and relationships between them

The distribution of these variables appears to deviate from a normal distribution.

HDI, Life expectancy, education exp, education mean is negatively correlated with GII, Mat mortality, adolescent birth.
GII and maternity mortality are also negatively correlated with education for female and male.

HDI, Life expectancy, education exp, education mean is positively correlated with education of male and female.
GII, adolescent birth and maternity mortality are also positively correlated.

\> This observed correlations appears coherent, highlighting logical relationships between the various variables.

## Question 2

Perform principal component analysis (PCA) on the raw (non-standardized) human data.
Show the variability captured by the principal components.
Draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables.
(0-2 points)

### Non-standardized PCA

```{r fig.height=20, fig.width=20}

# PCA
pca_human <- prcomp(human_new)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2)
# GNI has too big variances and is dominating the pca. The magnitude is too different between the metrics
```

## Question 3 - standardized PCA

Standardize the variables in the human data and repeat the above analysis.
Interpret the results of both analysis (with and without standardizing).
Are the results different?
Why or why not?
Include captions (brief descriptions) in your plots where you describe the results by using not just your variable names, but the actual phenomena they relate to.
(0-4 points)

### Scaled data & standardized PCA

```{r fig.height=15, fig.width=15}

# standardization of human_new
scaled_human <- scale(human_new)

# new PCA with scaled data
pca_human2 <- prcomp(scaled_human)

# draw a biplot of the PCA, adding captions that describe the phenomena
biplot(pca_human2, choices = 1:2,
   xlab="PC1 - Country Socio-Economic Development", ylab="PC2 - Female Empowerment and Socio-Economic Development by Women", main = " Principal Component Analysis of Socio-Economic Indicators: Mapping Country Development Factors and Female Empowerment")

# Check for missing values
any(is.na(scaled_human))

# Check for infinite values
any(is.infinite(scaled_human))
```

### Include captions and descriptions on PCA:

PC1 - Country Socio-Economic Development.
This component shows multiple socio-economic factors contributing to overall country development.
It includes on one side, variables like Gross National Income, Human Development Index, Life Expectancy, Education, and on the other side, variables like Adolescent birth rate, Maternal mortality ratio.
This axis reflects a broad aspect of a country's overall progress and development.

PC2 - Female Empowerment and Socio-Economic Development by Women.
This component specifically highlights aspects related to female empowerment, participation, and their impact on socio-economic development.
It might incorporate variables such as Female Education, Female Labor Force Participation.

## Question 4 - interpret the PCA

Give your personal interpretations of the first two principal component dimensions based on the biplot drawn after PCA on the standardized human data.
(0-2 points):

In the first PCA, the GNI variable has too high metrics and too big variances vs the other variables.
This variable dominates the PCA and therefore the PCA cannot be analyzed properly.
The magnitude is too different between the metrics to compare them together without scaling them.
After scaling the dataset, I did the PCA again to compare them more equally.

Analyze the standardized PCA: The arrows aligned with PC1 and its associated variables provide insight into the relationships between the variables themselves, it is the same for the PC2.
The direction of the arrow suggests that the pointed variables have a significant and positive influence on the principal components represented in the plot.
The direction of the arrow indicates which way the variable is correlated with the principal component.
The length of the arrow represents the strength of the variable's contribution to that principal component.
Longer arrows indicate a stronger influence on that component.

On the left side of the chart, countries appear more developed, characterized by higher values in key development indicators such as Gross National Income, Life Expectancy at Birth, Human Development Index, Life Expectancy in Education, and Female Population with Secondary Education.
These variables has significant influence, suggesting that the variables impact positively each other.
As these metrics increase, the level of development tends to be higher.
We can find in this area many European/Western countries.
Some of them also have the characteristics of higher labor rate among females and a closer rate when compared to the labor rate for males (e.g Iceland, Norway, Denmark).
Conversely, countries like Qatar, Bahrain, UAE, and Saudi Arabia display a relatively good development level but possess lower female labor participation, impacting the overall women development and freedom.

On the opposite side of the chart, variables such as Adolescent Birth Rate, Maternal Mortality Ratio, Male Labor Force Participation Rate, HDI Rank, and GNI Rank (where higher ranks denote lower HDI or GNI) portray countries facing more challenging situations.
Higher values in these variables correlate with less favorable conditions in a country.
As these indicators rise, the situation tends to be less favorable.
These countries, primarily from the Global South, such as Burkina Faso, Haiti, Niger, and Sierra Leone, face lower overall development.
Women situations also vary based on their positioning along PC2, reflecting the diverse and inequal experiences of women in different countries.

## Question 5

### Tea data import and exploration

The tea data comes from the FactoMineR package and it is measured with a questionnaire on tea: 300 individuals were asked how they drink tea (18 questions) and what are their product's perception (12 questions).

In addition, some personal details were asked (4 questions).
Load the tea dataset and convert its character variables to factors: tea \<- read.csv("<https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv>", stringsAsFactors = TRUE)

Explore the data briefly: look at the structure and the dimensions of the data.
Use View(tea) to browse its contents, and visualize the data.

```{r fig.height=30, fig.width=30}
tea_new <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)

# View and summary of the dataset
view(tea_new)
str(tea_new)
summary(tea_new)
# variables are factorial, except few ones. let's map the different categories thanks to the MCA

# But first let's keep only the categorical variables. An alternative would have been to change the non-catgeorical variables to categorical ones.
tea_new_2 <- tea_new %>% dplyr::select(-breakfast,-age)

```

### Multiple Correspondence Analysis (MCA)

On the tea data (or on just certain columns of the data, it is up to you!).
Interpret the results of the MCA and draw at least the variable biplot of the analysis.
You can also explore other plotting options for MCA.
Comment on the output of the plots.
(0-4 points)

```{r fig.height=10, fig.width=10}

library(FactoMineR)

# MCA
mca <- MCA(tea_new_2, graph = FALSE)
mca

# summary of the MCA
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), graph.type = "classic",col.var = rainbow(ncol(tea_new_2)))
# It allows us to visualize many different categories from many catgeorical variables, all at once to see the categories that fit together and suggest association (similarities in behavior). For instance, unpackages and tea shop seem close to each other and separated from the rest of the categories. It suggests that they share some commonality / similar patterns or behavior in term of users preferences. Besides, they tend to occur together within the dataset.

# Select smaller dataset to do the MCA analyze with less variables
# let's now choose only few variables that could have correlations together but also show distinct behavior between users. 
tea_new_2 %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = value)) +
  facet_wrap(~ name, scales = "free") +
  geom_bar() +
  labs(title = "Distribution of Categorical Variables") + 
  theme(axis.text.x = element_text(angle = 25, hjust = 1, size = 10)) +
  theme(strip.text = element_text(size = 10))
 
# smaller data set - I will look at 2 different ones. 
tea_smaller <- tea_new %>% dplyr::select(dinner, effect.on.health, frequency, evening, home, healthy, how, lunch, pub, price,Tea, where, work, sex)

tea_smaller2 <- tea_new %>% dplyr::select(diuretic, feminine, healthy, exciting, iron.absorption, escape.exoticism, relaxing, sophisticated, sex, slimming)

# MCA models
mca2 <- MCA(tea_smaller, graph = FALSE)
mca3 <- MCA(tea_smaller2, graph = FALSE)

# summary of the models
summary(mca2)
summary(mca3)

# visualize MCA 
plot(mca2, invisible=c("ind"), graph.type = "classic")
plot(mca3, invisible=c("ind"), graph.type = "classic")

```

### Conclusion of MCAs

MCA 2: The categories "tea shop," "upscale price," and "unpackaged" appear closely related, indicating they collectively represent a particular type of behavior or consumer preference.
Similarly, "tea bag" and "chain store" seem to indicate anothe pattern within the dataset.

MCA 3: Categories associated with tea benefits like "iron absorption," "healthy," "sophisticated," and "diuretic" form a cluster, suggesting they share similar characteristics or appeal to a specific segment of consumers.
Oppositely, tea without special benefits stands apart, indicating a different category or behavior.
Moreover, the two categories "exciting" and "not relaxing" but also the two categories "feminine tea" and "female" show similarities, aligning with their expected conceptual correspondence in this context.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
