---
editor_options: 
  markdown: 
    wrap: 72
---

# About the project

[My git repository](https://github.com/margot-dbz/IODS-project)

```{r}
date()
```

**Heads-up:** I am new with R. During this first week, I studied the 3
first chapters of the book made on R studio and watched the given
tutorial videos.\
\
**I have learned how to:**\
- set my directory\
- use the R project mode, the R markdown file and github commit\
- install and use libraries\
- identify and use the different data types including the factors \
- use the pipe %\>% (it means "and then...") \
- open data set, read data set, read the structure of the data set \
- select rows, columns and match text or rename column name\
- filter data based on variable values\
- create a table / tibble \
- create an object that is reusable later\
- use the time format\
- add columns with text, int, calculations, categories etc.\
- group data with one or many variables to then perform calculation\
- combine all data types to create a string with the function paste()\
- check the NA and remove the NA to do calculations \
- summarize to sum up data per variable value\
- manipulate table from wide to long and long to wide\
- combine 2 tables (functions: bind, inner_join, full_join, left_join,
right_join)\
- use the most common arithmetic functions and use percentage\
- rank and reorder value

With this tool I discovered how easy and fast it can be to manipulate
and visualize data from a big data set.\
\
**Conclusion of Week 1:\
**The book is really well made. The style is easy to follow and
understand. We can learn at our own pace and try the examples in R
studio to assimilate the concepts well.\
I am new with R and I find it hard to learn so many new different
concepts in a week. I would have liked to have more exercises, without
the solutions directly written under and with a tool to correct our
lines of code. Finally, I understand it is not necessary to know by
heart every function, but I would like to understand well the
capabilities of this tool by practicing and doing more real life
exercises.\
\
**Lines of code to summarize learning in week 1:**

```{r}

# Get the data set
library(tidyverse)
library(scales)
gbd_full <- read_csv("https://raw.githubusercontent.com/KimmoVehkalahti/RHDS/master/data/global_burden_disease_cause-year-sex-income.csv")

# This is a code I wrote to know the pourcentage of deaths per cause for each year.
exercise4 <- gbd_full %>% 
  group_by(year, cause) %>% 
  summarise(total_per_cause= sum(deaths_millions)) %>% 
  group_by(year) %>% 
  mutate(total_per_year =sum(total_per_cause)) %>% 
  mutate(percentage = percent(total_per_cause/total_per_year)) %>% 
  select(year,cause,percentage) %>% 
  pivot_wider(names_from = cause, values_from = percentage)
  

exercise4

# This is a code I wrote to know the number of deaths per sex and income and see the ratio Male / Female for each income type.
gbd_full %>% 
  filter(year ==1990) %>% 
  group_by(sex,income) %>% 
  summarize(deaths_per_sex_income = sum(deaths_millions)) %>% 
  pivot_wider(names_from = sex, values_from = deaths_per_sex_income) %>% 
  mutate(diff_M_F = Male/Female)

```
