---
editor_options: 
  markdown: 
    wrap: 72
---

# Chapter 3

```{r}

setwd("/Users/margot/Desktop/Desktop - MacBook Pro de MARGOT/Open data with R 2023/IODS-project")

library(broom)
library(ggplot2)
library(dplyr)
data <- read.csv("alc.csv",sep = ",", header = T)

class(data$high_use)
class(data$famrel)
class(data$goout)
class(data$studytime)

```

## Question 3, 4 and 5

The purpose of your analysis is to study the relationships between
high/low alcohol consumption and some of the other variables in the
data. To do this, choose 4 interesting variables in the data and for
each of them, present your personal hypothesis about their relationships
with alcohol consumption. (0-1 point)

### Hypothesis 1: students with a good family relationship have less chance to have a high consumption of alcohol

Comments on the below codes chunk: There is a negative correlation
between family relationship and the consumption of alcohol. When the
relationship is not good (from 1 to 5), there are more risk to drink lot
of alcohol. Pvalue is 0,0204 which is representative(\<0.05). The
coefficient for the variable Famrel in this model is -0.2838. The odds
ratio associated with a one-unit change in Famrel is approximately
exp(-0.2838) = 0.753. For a one-unit decrease in Famrel, the odds of
'high_use' decrease by about 1 - 0.753 = 24.7%. With 95% confidence, the
true value of the odds ratio for the intercept lies between
approximately 0.49 and 3.35. For the famrel, the odds ratio lies between
0.59 and 0.95, since it is less than 1, there is a statistical
significance, with a negative effect from the variable famrel to the
high_use. The null deviance is bigger than the residual deviance of the
model. It shows that the model explains a part of the variable high-use.

```{r}

# I add the family = binomial as we do the analyze on a binary variable (True and False)
logistic_model_family_alcohol <-  glm(high_use ~ famrel, data = data, family = binomial)
summary(logistic_model_family_alcohol)


# get the coef
coef(logistic_model_family_alcohol) %>% exp()

# odd of being in high use - exponential of the coef
exp(coef(logistic_model_family_alcohol))

# get the intervals
confint(logistic_model_family_alcohol) %>% exp()

# get the variance, BIC and AIC
logistic_model_family_alcohol %>% 
  glance()

# If I want to check the impact of high use to family relationship -> famrel is ordinal so I use the ordinal regression model
library(ordinal)
ordinal_model_family_alcohol <- clm(factor(famrel) ~ high_use, data = data)
summary(ordinal_model_family_alcohol)

# This box plot confirms our hypothesis and conclusion. We can see that the mean and median of the family relationship quality is very different whether the students drink a lot or not. 
data %>% 
  ggplot(aes(x= factor(high_use), y= famrel)) +
  geom_boxplot() +
  labs(title = "Quality of family relationship vs the alcohol consumption",
       x = "High consumption",
       y = "quality of family relationship")

data %>%
  ggplot(aes(x = as.factor(famrel), fill = as.factor(high_use))) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Family relationship and Alcohol Consumption",
    x = "Family relation score from 1 to 5",
    y = "Proportion of students",
    fill = "High Use of alcohol: True of False"
  )

# there are less score 5 and way more score 2 among the students who drink a lot
data %>%
  ggplot(aes(x = as.factor(high_use), fill = as.factor(famrel))) +
  geom_bar(position = "fill") +
  labs(
    title = "Family relation score distribution by Alcohol Consumption",
    x = "High Use of alcohol: True of False",
    y = "Proportion of students",
    fill = "Family relation score from 1 to 5"
  )



```

### Hypothesis 2: students who go more out with their friends have more risk to have a high consumption of alcohol

Comments on the below codes chunk: There is a a positive relationship
between going out and the consumption of alcohol and a significant p
value (very small value), which confirms the hypothesis. The more
students go out the more risks they have to have a high consumption in
alcohol.

The coefficient for goout is 0.7563.The odds ratio associated with a
one-unit change in Goout is approximately exp(0.7563) = 2.13.For a
one-unit increase in Goout, the odds of 'high_use' increase by about
2.13 - 1 = 1.13, which is about 113% increase.

With 95% confidence, the true value of the odds ratio for the intercept
lies between approximately 0.015 and 0.078. For the goout, the odds
ratio lies between 1.71 and 2.66, since it is \> 1, there is a
statistical significance with a positive effect of this variable on the
high_use.

The null deviance is way bigger than the residual deviance of the model.
It shows that the model explains the variable high-use, even better than
in the hypothesis 1. The AIC of this model is even smaller than the
previous model too, which confirms my conclusion.

```{r}
logistic_model_goout_alcohol <- glm(high_use ~ goout, data = data, family = binomial)
summary(logistic_model_goout_alcohol)
tidy(logistic_model_goout_alcohol)

# get the coef
coef(logistic_model_goout_alcohol) %>% exp()

# get the exp of the coef
exp(coef(logistic_model_goout_alcohol))

# get the intervals
confint(logistic_model_goout_alcohol) %>% exp()

# plots
data %>% 
  ggplot(aes(x= factor(high_use), y= goout)) +
  geom_boxplot() +
  labs(title = "Going out with friends vs the alcohol consumption",
       x = "High consumption",
       y = "go out with friends (from 1 to 5)")


# the students who are drinking more are more numerous among the students who gave the score 4 and 5 for the goout variable.
data %>%
  ggplot(aes(x = factor(goout), fill = factor(high_use))) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Going Out Behavior and Alcohol Consumption",
    x = "going out with friends from 1 to 5",
    y = "Proportion of students",
    fill = "High Use of alcohol: True of False"
  )

# there are more score 1, 2 and 3 among the students who do not drink a lot
data %>%
  ggplot(aes(x = factor(high_use), fill = factor(goout))) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Alcohol Consumption and going out behavior",
    x = "High Use of alcohol: True of False",
    y = "Proportion of students",
    fill = "going out with friends from 1 to 5"
  )

```

### Hypothesis 3: students who study a lot have less risk to have a high consumption of alcohol

Comments about the below code chunk: The p value is very small and
estimate is negative. We can conclude there is a negative relation
between the study time and the alcohol consumption. When the study time
increases the odds of high alcohol consumption decreases.

The coefficient for Studytime is -0.6208. The odds ratio associated with
a one-unit change in Studytime is approximately exp(-0.6208) = 0.538.
For a one-unit decrease in Studytime, the odds of 'high_use' decrease by
about 1 - 0.538 = 46.2%.

With 95% confidence, the true value of the odds ratio for the intercept
lies between approximately 0.79 and 2.67. For the study time, the odds
ratio lies between 0.39 and 0.72, since it is \< 1, there is a
statistical significance with a negative effect of this variable on the
high_use.

The residual deviance is smaller than the null deviance, it means the
model explains the variable high use. The AIC is smaller than the first
model but not as small as the second model.

```{r}
logistic_model_studytime_alcohol <- glm(high_use ~ studytime, data = data, family = binomial)
summary(logistic_model_studytime_alcohol)
tidy(logistic_model_studytime_alcohol)

# get the coef
coef(logistic_model_studytime_alcohol) %>% exp()

#Get the odds - exponential of coefficient: 
exp(coef(logistic_model_studytime_alcohol))

# get the intervals
confint(logistic_model_studytime_alcohol) %>% exp()

# This plot shows the difference in median and mean for the students who drink a lot or not
data %>% 
  ggplot(aes(x= factor(high_use), y= studytime)) +
  geom_boxplot() +
  labs(title = "Study time vs the alcohol consumption",
       x = "High consumption",
       y = "Study time (from 1 to 4)")

# way more students among students who score they study time around 1 and 2
data %>%
  ggplot(aes(x = factor(studytime), fill = factor(high_use))) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Study time Behavior and Alcohol Consumption",
    x = "Study time from 1 to 4",
    y = "Proportion of students",
    fill = "High Use of alcohol: True of False"
  )

# way more score 1 and way less 3 and 4 among students who drink a lot.
data %>%
  ggplot(aes(x = factor(high_use), fill = factor(studytime))) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Alcohol Consumption and Study time behavior",
    x = "High Use of alcohol: True of False",
    y = "Proportion of students",
    fill = "Study time from 1 to 4"
  )
```

### Hypothesis 4: students who chose the school because of its reputation has less risk to have a high consumption of alcohol

Comments about the below code: The p value is 0,022 which tells that
there is a correlation between the reason to choose a school and the
consumption of alcohol.

We don't know with the test which reason is the most chosen by the
students who drink a lot of alcohol vs the ones who don't. Some plots
can help us identify it (check below).

```{r}


# Creating a contingency table
chi_table_alcohol_reasons <- table((data$high_use), as.factor(data$reason))

# Performing the chi-squared test
chi_square_test <- chisq.test(chi_table_alcohol_reasons )
chi_square_test

# Summary of the test
summary(chi_square_test)

# This show that students who don't drink a lot have chosen their school more for its reputation than the one who drink a lot. For the ones who drink, the reasons home and other are way more important than for the students who don't drink too much. The reason courses seem to be equally important for them.
data %>%
  ggplot(aes(x = high_use, fill = reason)) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Reason to chose their school and Alcohol Consumption",
    x = "High Consumption of a",
    y = "Proportion of students",
    fill = "High Use of alcohol: True of False"
  )

data %>%
  ggplot(aes(x = reason, fill = high_use)) +
  geom_bar(position = "fill") +
  labs(
    title = "Relationship Between Alcohol Consumption and Reason to chose their school",
    x = "High Use of alcohol: True of False",
    y = "Proportion of students",
    fill = "Reason to choose the school"
  )
```

### Other models to test:

By adding each variable together, we can see that the variation in high
use is better explained (this is because the deviance is decreasing from
452 to 379).

All three variables (goout, famrel, and studytime) have statistical
significance in predicting alcohol consumption (high_use).

```{r}
logistic_model_3var <- glm(high_use ~ goout + famrel + studytime, data = data, family = binomial)
logistic_model_3var_2 <- glm(high_use ~ goout * famrel * studytime, data = data, family = binomial)
summary(logistic_model_3var)
tidy(logistic_model_3var)

# Anova conclusion: The p-value associated with the difference in deviance (Pr(>Chi)) is 0.1362, suggesting that the difference in fit between the two models is not statistically significant
anova(logistic_model_3var, logistic_model_3var_2, test="LRT")

# when comparing with for instanc: logistic_model_family_alcohol, the residual deviance of logistic_model_3var is better and the p is significant. 
anova(logistic_model_3var, logistic_model_family_alcohol, test="LRT")

# best model is the logistic_model_3var according to BIC and AIC, because metrics are smaller for this model
AIC(logistic_model_3var, logistic_model_3var_2)
BIC(logistic_model_3var, logistic_model_3var_2)


```

## question 6: Model 1 (high_use \~ goout + famrel + studytime)

Using the variables which, according to your logistic regression model,
had a statistical relationship with high/low alcohol consumption,
explore the predictive power of you model. Provide a 2x2 cross
tabulation of predictions versus the actual values and optionally
display a graphic visualizing both the actual values and the
predictions. Compute the total proportion of inaccurately classified
individuals (= the training error) and comment on all the results.
Compare the performance of the model with performance achieved by some
simple guessing strategy.

```{r}
# prediction based on the model: logistic_model_3var
predictions <-  predict(logistic_model_3var, type = "response")

#Create the confusion matrix: 
table(data$high_use, predictions > 0.5)
# correct are False false + true true = 235 + 48 = 293
# wrong are False True and True False = 24 + 63 = 87

# Probability of miss classifications  = 87/(293+87) = 23% 
# odds of miss classification = 87/293 = 30%

# precision: correctly positive among all positives predictions = 0.33%
precision <- 24 / (24 + 48)
# recall: correctly positive among all actual positives = 0.28%
recall <- 24 / (24 + 63)
# F1 score: 0.30%
 2 * (precision * recall) / (precision + recall)

 # Accuracy of always predicting the majority class:  if the majority class is "FALSE," then predicting "FALSE" for every instance would result in a correct prediction rate of 70%
mean(data$high_use == FALSE)

# Calculate the total proportion of errors is 23.5%. This means my model is a bit better than the majority class which is 70% (30% of non-accuracy).
mean(data$high_use != (predictions > 0.5))


# to conclude, my model is a bit better than using the prediction based on the majority class. But it should be improved for better accuracy!
```

### Bonus:

I had to look at chat gpt fully for this one. I did not know the method.
I now understand this method splits the data in 10 data samples and will
use 9 of them to predict the last one, rotating to test them all. In
that case, the average of prediction for the different tests is 74.9%
which is better than the 70% (majority class).

```{r}
library(caret)
library(lattice)
# Define the model formula
formula <- factor(high_use) ~ goout + famrel + studytime

# Define the number of folds for cross-validation
num_folds <- 10

# Create a training control with 10-fold cross-validation
train_control <- trainControl(method = "cv", number = num_folds)

# Train the model using 10-fold cross-validation
model <- train(formula, data = data, method = "glm", family = binomial, trControl = train_control)

# Get cross-validated results
results <- model$resample
str(results)

# Calculate mean prediction error
mean(results$Accuracy)



```
